---
title: "Accessing CLAMPData ExperimentHub resources"
author: "Marc Subirana-Granés"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Accessing CLAMPData ExperimentHub resources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(CLAMPData)
library(ExperimentHub)
```

# Introduction

This vignette demonstrates how to access data resources provided by **CLAMPData**, the companion ExperimentHub data package for **CLAMP**, through the [ExperimentHub](https://bioconductor.org/packages/release/bioc/html/ExperimentHub.html) interface.

CLAMPData provides curated gene-set libraries, pathway priors, and example expression matrices used by the CLAMP software package for prior-informed latent variable modeling.

# Available Resources

CLAMPData provides three ExperimentHub resources:

| Resource ID | Title | Description |
|-------------|-------|-------------|
| EH10279 | GSE164416_DP_htseq_counts_txt_gz | HTSeq gene counts for islet RNA-seq example |
| EH10280 | human_gene_v2_5_alz_h5 | HDF5 file with gene-set priors for CLAMP |
| EH10281 | islets_metadata_csv | Sample metadata for islet RNA-seq example |

# Accessing data resources

```{r list, message=FALSE, warning=FALSE}
p_counts <- GSE164416_DP_htseq_counts_txt_gz()
p_h5     <- human_gene_v2_5_alz_h5()
p_meta   <- islets_metadata_csv()

cat("Counts path:", p_counts, "\n")
cat("H5 path:", p_h5, "\n")
cat("Metadata path:", p_meta, "\n")
```

# Reading example data

```{r accessors, message=FALSE, warning=FALSE}
# --- Counts Data (GSE164416) ---
counts <- read.delim(gzfile(p_counts), row.names = 1)
cat("Counts matrix dimensions:", nrow(counts), "genes x", ncol(counts), "samples\n\n")
knitr::kable(counts[1:5, 1:5], caption = "Gene counts (first 5 genes × 5 samples)")

# --- HDF5 Gene-Set Priors ---
if (requireNamespace("rhdf5", quietly = TRUE)) {
  cat("\nHDF5 file structure:\n")
  knitr::kable(rhdf5::h5ls(p_h5))
} else {
  message("Install 'rhdf5' from Bioconductor to inspect HDF5 contents: ",
          "BiocManager::install('rhdf5')")
}

# --- Sample Metadata ---
meta <- read.csv(p_meta, stringsAsFactors = FALSE)
cat("\nMetadata dimensions:", nrow(meta), "samples x", ncol(meta), "variables\n")
cat("Sample types:", paste(unique(meta$type), collapse = ", "), "\n\n")
knitr::kable(head(meta), caption = "Sample metadata (first 6 rows)")
```

# Session info

```{r session}
sessionInfo()
```
